---
title: "Data Frames in R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<style> p { text-align: justify; } </style>

## **What is a Data Frame?**

Data frames are a cornerstone of data manipulation and analysis in R, designed for handling tabular data efficiently. They are two-dimensional, rectangular structures where rows represent observations and columns represent variables. Each column is a vector of equal length, allowing for the storage of mixed data types such as numerical, character, or logical values. Analogous to spreadsheets or SQL tables, data frames are indispensable in tasks like statistical analysis, data visualization, and machine learning.

Key characteristics of data frames:

- Columns can contain different data types (e.g., numeric, character, factor).
- Rows represent individual observations or cases.
- Data frames can hold large datasets efficiently.

## **Creating Data Frames**

We can create a data frame in R using the `data.frame()` function from base R. Here's an example:

```{r create_df, message=FALSE, warning=FALSE, include=TRUE}

# Creating a data frame
students <- data.frame(
  Name = c("Amrit", "Ishwor", "Indra", "Rakshya"),
  Age = c(27, 24, 21, 25),
  Grade = c("A", "B", "A", "B"),
  Major = c("Maths", "Science", "Maths", "Science")
)

# Display the data frame
print(students)

```

## **Accessing and Modifying Data**

#### **Selecting Rows and Columns**

We can access elements of a data frame using indexing, column names, or logical conditions.

```{r access, message=FALSE, warning=FALSE, include=TRUE}

# Access a specific column
students$Name

# Access a specific row
students[1, ]

# Access a specific value
students[2, "Grade"]

```

#### **Adding and Removing Columns**

We can add a new column to a data frame dynamically or remove an existing column.

```{r modify, message=FALSE, warning=FALSE, include=TRUE}

# Adding a new column
students$Passed <- c(TRUE, FALSE, TRUE, FALSE)
# Display the data frame
print(students)

# Removing a column
students$Grade <- NULL
# Display the data frame
print(students)

```

## **Data Frame Operations**

#### **Filtering and Subsetting**

We can filter rows based on conditions using logical operators.

```{r filter, message=FALSE, warning=FALSE, include=TRUE}

# Filter students with Age > 22
students[students$Age > 22, ]
```

#### **Sorting and Reordering**

Sorting a data frame is straightforward with the `order()` function.

```{r reorder, message=FALSE, warning=FALSE, include=TRUE}

# Sort by Age
students <- students[order(students$Age), ]
# Display the data frame
print(students)

```

#### **Summary and Statistical Functions**

R provides built-in functions to summarize and analyze data frames.

```{r summary, message=FALSE, warning=FALSE, include=TRUE}

# Summary statistics
summary(students)

# Number of rows and columns
nrow(students)
ncol(students)

```

#### **Aggregation and Grouping**

The `dplyr` package functions enable grouping and summarization.

```{r group, message=FALSE, warning=FALSE, include=TRUE}

# Load dplyr package
library(dplyr)

# grouping
students %>%
  group_by(Major) %>%
  summarise(Average_Age = mean(Age))

```

So far, this chapter has introduced you to the versatility and power of data frames in R, laying the foundation for data manipulation and analysis. Now, we will delve into data frames specifically designed for geospatial data.

























