---
title: "Functions in R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<style> p { text-align: justify; } </style>

Functions are blocks of code that perform a specific task in R. They allow us to encapsulate logic and reuse it across different parts of the code, making our scripts more modular, efficient, and easy to debug. Functions are one of the most powerful tools in R and are commonly used in data analysis, machine learning, and statistical programming.

By using functions, we can:

- Avoid repetitive code
- Break complex tasks into smaller, manageable pieces
- Improve readability and maintainability of your code

## **Creating Functions in R**

In R, a function is created using the **function()** keyword. The basic syntax for defining a function is:

```{r syntax, message=FALSE, warning=FALSE, include=TRUE}

my_function <- function(arg1, arg2) {
  # function body
  result <- arg1 + arg2
  return(result)
}
```

Here’s what each part means:

- `my_function`: The name of the function.
- `function(arg1, arg2)`: The definition of the function with its arguments.
- `{}`: The body of the function where the logic is written.
- `return(result)`: The value that is returned by the function.

Example: 

```{r example1, message=FALSE, warning=FALSE, include=TRUE}

# A simple function to add two numbers
add_numbers <- function(a, b) {
  sum <- a + b
  return(sum)
}

# Call the function
add_numbers(5, 3)
```

## **Function with Multiple Arguments**

Functions can have multiple arguments, and you can pass values in the form of position or by explicitly naming the arguments when calling the function.

```{r example2, message=FALSE, warning=FALSE, include=TRUE}

# A function to calculate the area of a rectangle
rectangle_area <- function(length, width = 2) {  # Default value for width
  area <- length * width
  return(area)
}

# Call the function with default width
rectangle_area(5)

# Call the function with a specific width
rectangle_area(5, 3)
```

In the above example, the argument width has a default value of 2. If you don't provide a value, R will use this default.






